Closing the gap between statistical and scientific workflows for improved forecasts in ecology

### 2025-03-26 forecastflows.tex
adding some refs

\documentclass[11pt]{article}
\usepackage[top=1.00in, bottom=1.0in, left=1in, right=1in]{geometry}
\renewcommand{\baselinestretch}{1.1}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{amsmath}

\begin{document}
\renewcommand{\refname}{\CHead{}}

@@

% Ecology has become super challenged to predict stuff for decision making
Nature is increasingly threatened by multiple drivers of change, with a largely dominant influence of human activities. This ongoing biodiversity crisis is expected to increase in the next decades, and will continue to alter ecosystem services and human well-being. To support implementation of sustainable policies among the socioeconomic and environmental dimensions, it is critical to understand trends to date and be able to forecast future dynamics. 
% $\rightarrow$ Example of stuff to predict: population size and geographic distribution (terrestrial, freshwater, marine), extinction risk,...

% General way to do this so far... (bifurcated?)
Estimation of global biodiversity indicators and current trends depends on large-scale and long-term datasets---across terrestrial, freshwater, and marine ecosystems. These data, gathered opportunistically and from multiple sources, are often unbalanced and have geographic, temporal and taxonomic biases. Addressing these biases requires the use of appropriate statistical inference.
Forecasting future changes---under different plausible scenarios---generally relies on either correlative models or process-based models. The latter, which focus on a mechanistic representation of ecosystem functioning, are often promoted as the most realistic approach.

% Gap, problem: None of this is going well
The urgent need to answer policy-relevant questions has favored the proliferation of diverse methods developed by different researchers, lacking an overall coherence. Though there is no doubt nature is declining globally, significant uncertainty remains. There is no consensus on current species trends, with ongoing debates driven by widely varying reports that sometimes show conflicting trend directions. Future projections also diverge considerably, due to a high model uncertainty at the ecological level. Predictive modeling is increasingly relying on overly complex models (with a huge number of parameters), making it less adequate to generate new scientific insights.

% Current workflows in ecology are not up to the task
% Here we introduce a universal workflow to adress this...
Ongoing controversies and focus on methodological aspects stem from the lack of coherence between current workflows in ecology. Each new model development is added as a separate layer, disconnected from the original research aim, the data stream and the previous scientific insights. Workflows should fully integrate all the steps required to build a model from an ecological question, evaluate its limitations and degeneracies, before estimating its parameters and making projections. Here, we introduce an universal workflow that proposes to iteratively build upon all these steps, harmonizing both trend estimation and forecasting, with the aim of refocusing the debate on ecological questions.

@@

@@

\begin{enumerate}
\item
\end{enumerate}

\citep{ospreebbms}

\bibliographystyle{/Users/Lizzie/Documents/EndnoteRelated/Bibtex/styles/besjournals}
\bibliography{/Users/Lizzie/Documents/git/bibtex/LizzieMainMinimal}

\clearpage
\section{Figures}

\begin{figure}[h!]
\includegraphics[width=0.4\textwidth]{..//..//..//..//Professional/images/ColylusVineFruit2.png}
\caption{I am a caption.} 
\label{fig:figname}
\end{figure}

>>>>

\documentclass[11pt]{article}
\usepackage[top=1.00in, bottom=1.0in, left=1in, right=1in]{geometry}
\renewcommand{\baselinestretch}{1.1}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{amsmath}
\bibliographystyle{besjournals}

\begin{document}

@@

% Ecology has become super challenged to predict stuff for decision making
Nature is increasingly threatened by multiple drivers of change, with a largely dominant influence of human activities \citep{Diaz2019}. This ongoing biodiversity crisis is expected to increase in the next decades because of climate change, and will continue to alter ecosystem services and human well-being \citep{IPBES2019}. To support implementation of sustainable policies among the socioeconomic and environmental dimensions, it is critical to understand trends to date and be able to forecast future dynamics. 
% $\rightarrow$ Example of stuff to predict: population size and geographic distribution (terrestrial, freshwater, marine), extinction risk,...

% General way to do this so far... (bifurcated?)
Estimation of global biodiversity indicators and current trends depends on large-scale and long-term datasets---across terrestrial, freshwater, and marine ecosystems \citep[e.g.][]{Dornelas2018}. These data, gathered opportunistically and from multiple sources, are often unbalanced and have geographic, temporal and taxonomic biases. Addressing these biases requires the use of appropriate statistical inference.
Forecasting future changes---under different plausible scenarios---generally relies on either correlative models or process-based models \citep{IPBES2019}. The latter, which focus on a mechanistic representation of ecosystem functioning, are often promoted as the most realistic approach \citep{Urban2016, Pilowsky2022}. % ... lacks something

% Gap, problem: None of this is going well
The urgent need to answer policy-relevant questions has favored the proliferation of diverse methods developed by different researchers, lacking an overall coherence. Though there is no doubt nature is declining globally, significant uncertainty remains. There is no consensus on current species trends, with ongoing debates driven by widely varying reports that sometimes show conflicting trend directions \citep{Dornelas2014, Leung2020, Buschke2021, Johnson2024}. Future projections also diverge considerably, due to a high model uncertainty at the ecological level. Predictive modeling is increasingly relying on overly complex models (with a huge number of parameters), making it less adequate to generate new scientific insights \citep{Franklin2020}.

% Current workflows in ecology are not up to the task
% Here we introduce a universal workflow to adress this...
Current controversies and focus on methodological aspects stem from the lack of coherence between current workflows in ecology \citep{Loreau2022, Talis2023, Johnson2024}. Each new model development is added as a separate layer, disconnected from the original research aim, the data stream and the previous scientific insights. Workflows should fully integrate all the steps required to build a model from an ecological question, evaluate its limitations and degeneracies, before estimating its parameters and making projections. Here, we introduce an universal workflow that proposes to iteratively build upon all these steps, harmonizing both trend estimation and forecasting, with the aim of refocusing the debate on ecological questions and increasing the speed of scientific progress.

@@

\clearpage
\bibliography{forecastflows}

@@
